Converting LaTeX R Journal Articles into Rmarkdown using texor and rebib

### 2023-07-19 texor.tex
Included rendered html,pdf versions of the article

% !TeX root = RJwrapper.tex
\title{Converting `LaTeX' `R Journal' Articles into `RJ-web-articles'}
\author{by Abhishek Ulayil}

@@

Explain the purpose of the work, some background on the rjournal/rjtools{[}DC{]},

\hypertarget{background}{%
\section{Background}\label{background}}

LaTeX, Regex, pandoc
building blocks of the packages, methods used in them {[}MOW{]}

\hypertarget{texor-package-converting-latex-to-rmarkdown}{%
\section{texor package converting LaTeX to Rmarkdown}\label{texor-package-converting-latex-to-rmarkdown}}

\hypertarget{also-mention-the-hard-parts}{%
\subsection{Also mention the hard parts}\label{also-mention-the-hard-parts}}

\hypertarget{lua-filters-with-pandoc}{%
\subsection{Lua Filters with pandoc}\label{lua-filters-with-pandoc}}

A lot of details about different components of texor

\hypertarget{rebib-package-working-out-bibliography}{%
\section{rebib package working out bibliography}\label{rebib-package-working-out-bibliography}}

\hypertarget{getting-started}{%
\section{getting started}\label{getting-started}}

Examples/ getting started / usage and so on

@@

\hypertarget{supplementary-materials}{%
\section{supplementary Materials}\label{supplementary-materials}}

github repos for the packages, rjtools

@@

>>>>

% !TeX root = RJwrapper.tex
\title{Converting LaTeX R Journal Articles into Rmarkdown using texor and rebib}

\author{by Abhishek Ulayil, Mitch O'Hara-Wild, Christophe Dervieux, Heather Turner, and Dianne Cook}

@@

The R Journal is the primary open-access outlet for publications produced by the R community. It was born in 2008, evolving from a newsletter, that ran from 2001, into a more formal article publication to encourage documenting statistical computing research.

The format is constantly evolving. Early articles were typeset using LateX (The LaTeX Project 2023), from a specific, but changing, template. This requires that code is separated from the documentation, and there is a chance that code chunks in the paper don't reproduce the results reported. With the emergence of dynamic document systems like R Markdown (Xie, Allaire, and Grolemund 2018) a tight-coupling of code and documentation is possible. Code chunks are dynamically executed when the document is typeset using a system like \CRANpkg{knitr} (Xie 2015), making reporting of computing research more reproducible.

In 2019, with the help of funding from the R Consortium it was decided that it was time to update operations. One aspect of this was to change from LaTeX paper submissions to a more reproducible format, where code was embedded in the document, and the output could be both HTML and pdf. There are numerous benefits of HTML format:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Articles can include interactive graphics and tables.
\item
  The format is more accessible to screen readers making the work more accessible to vision-impaired researchers.
\end{enumerate}

This latter point is a reason to consider converting all of the legacy articles into HTML.

A key decision for creating conversion software was to decide to directly convert LaTeX to HTML, or PDF to HTML, or LaTeX to R Markdown, and then use the current journal tools to create the HTML. The latter approach was decided to be the most versatile and useful. If an article can be converted from LaTeX to R Markdown, it would help authors make the transition to reproducible publishing, beyond what the R Journal needed. Once an article is in R Markdown format it can be adapted to include the code for dynamic execution.

In addition to article format, changes to the web site structure were important for delivering the publication. Web site architectures are also constantly evolving, and the emergence of \CRANpkg{distill} (Dervieux et al. 2022) allows for the journal web site to optimally deliver R Markdown articles.

The \CRANpkg{rjtools} was developed to create articles using R Markdown for the R Journal, and to embed them into the journal web site. The packages described here, \CRANpkg{texor} and \CRANpkg{rebib} describes software to convert legacy LaTeX format articles into Rmarkdown, so that they can be rendered in HTML in the new web site.

The paper is organised as follows. Section XXX gives an overview of the conversion process. Section XXX describes examples of pre-processing using regular expressions, and section XXX provides examples of post-processing using lua filters. Section XXX describes tools to do special handling of bibliography files. Section XXX runs through an example conversion process.

\hypertarget{converting-from-latex-to-r-markdown}{%
\section{Converting from LaTeX to R Markdown}\label{converting-from-latex-to-r-markdown}}

The decision to convert to R Markdown format means that the output to pdf and HTML will depend on Pandoc (MacFarlane 2023).
Pandoc is a versatile document conversion program written in Haskell that is core to numerous documentation systems, including Rmarkdown and Quarto. Pandoc first converts a document into an abstract syntax tree. From this, it can convert to a different format, including custom ones. Although creating bespoke writers/readers in \texttt{Lua} can be challenging, pandoc includes filters that allow users to customize the output generated based on matching patterns plus some logic to modify them.

\textbf{DC: Can you explain what a Lua filter is?}

The \CRANpkg{texor} package uses these filters heavily, even for simple tasks like choosing which supporting figures to copy. A \texttt{Lua} filter constructs a list of image paths, stores them momentarily and the package uses this data to copy images. However, \CRANpkg{texor} also includes many pre-processing functions that employ regular expressions. This is needed to convert some text and LaTeX commands that are not recognized or handled by Pandoc.

\textbf{DC: Add a diagram here showing flow of work, just pre-process .tex -\textgreater{} pandoc -\textgreater{} lua ???}

\textbf{Maybe more details on the supporting packages here: LaTeX, Regex, pandoc building blocks of the packages, methods used in them {[}MOW{]}}

\hypertarget{pre-processing-using-regular-expressions}{%
\subsection{Pre-processing using regular expressions}\label{pre-processing-using-regular-expressions}}

LaTeX is very descriptive and allows unfettered levels of customization to authors, which poses an issue during conversions. RMarkdown is quite restrictive by syntax though you can always embed HTML to expand the feature set. Although
pandoc does most of the heavy lifting, there are some instances where even filters cannot assist. A good demonstration of the issue in hand is code environments.

A limitation of pandoc is that it only works with a ``verbatim'' environment. However, in ``RJournal'', we have multiple custom code environments which are not supported or defined due to lower-level tex limitations in pandoc. So the code environments are eliminated during the conversion process. Including a Lua filter to find these elements and redefine them as verbatim environments would also be ineffective. If the code environment is not verbatim, then pandoc will also try to process the actual code content as LaTeX commands and will lose out on details in some cases.

\textbf{Be sure to mention the hard parts}

\hypertarget{post-processing-lua-filters}{%
\subsection{Post-Processing : Lua Filters}\label{post-processing-lua-filters}}

This was one of the earliest and most basic Lua filters, and it was part of the \CRANpkg{texor} package:

\begin{verbatim}
function Div(el)
    if el.classes[1] == 'thebibliography' then
        return { }
    end
end
\end{verbatim}

Above filter reads the abstract syntax tree and filters out all the Div elements. Then it looks for the class ``thebibliography.'' It turns out that this Div element contains the LaTeX bibliographic records, which usually appear at the very end
of papers. We don't need this portion in the article with the ``RJ-web-article'' layout, given its inclusion as meta-data in the footer.

\hypertarget{structure-of-the-package}{%
\section{\texorpdfstring{Structure of the \CRANpkg{texor} package}{Structure of the  package}}\label{structure-of-the-package}}

\textbf{DC: List the main functions in an organised way, could be the pre-processing functions, the post-processing functions, utility functions. This can be also be summarized in a table, but there needs to be text explanations of the main functionality.}

\hypertarget{using}{%
\section{\texorpdfstring{Using \CRANpkg{texor}}{Using }}\label{using}}

\textbf{Install instructions}

\textbf{Examples / getting started / usage and so on}

\hypertarget{managing-the-bibliography-using}{%
\section{\texorpdfstring{Managing the bibliography using \CRANpkg{rebib}}{Managing the bibliography using }}\label{managing-the-bibliography-using}}

\hypertarget{overview}{%
\subsection{Overview}\label{overview}}

The \CRANpkg{rebib} package addresses the issue with LaTeX articles using built-in bibliography options with or without BibTeX files. While this works well with LaTeX, it won't work with Rmarkdown. Initially the goal was to use external software like Biber to convert the embedded bibliography to BibTeX.
However the integration of external software was not viable, hence the experimental idea of a bibliography parser gained momentum.

It was initially a part of the \CRANpkg{texor} package, as those functions grew in features and became more involved; at that point, it made sense to move those functions as a separate package. Initially, there were some reservations about
the usage of \CRANpkg{rebib} and its stability with various formats. However, the package has improved over time and proven to be a good performer.

\hypertarget{using-1}{%
\subsection{\texorpdfstring{Using \CRANpkg{rebib}}{Using }}\label{using-1}}

\textbf{Install instructions}

\textbf{Examples / getting started / usage and so on}

@@

\hypertarget{supplementary-materials}{%
\section{Supplementary materials}\label{supplementary-materials}}

github repos for the packages, texor, rebib, rjtools

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-distill}{}}%
Dervieux, Christophe, JJ Allaire, Rich Iannone, Alison Presmanes Hill, and Yihui Xie. 2022. \emph{Distill: 'R Markdown' Format for Scientific and Technical Writing}. \url{https://CRAN.R-project.org/package=distill}.

\leavevmode\vadjust pre{\hypertarget{ref-pandoc}{}}%
MacFarlane, John. 2023. {``Pandoc: A Universal Document Converter.''} https://pandoc.org.

\leavevmode\vadjust pre{\hypertarget{ref-latex}{}}%
The LaTeX Project. 2023. {``LaTeX -- a Document Preparation System.''} https://www.latex-project.org.

\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
Xie, Yihui. 2015. \emph{Dynamic Documents with {R} and Knitr}. 2nd ed. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://yihui.org/knitr/}.

\leavevmode\vadjust pre{\hypertarget{ref-rmarkdown}{}}%
Xie, Yihui, J. J. Allaire, and Garrett Grolemund. 2018. \emph{R Markdown: The Definitive Guide}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown}.

\end{CSLReferences}

@@

\address{%
Mitch O'Hara-Wild\\
Monash University\\%
Melbourne, Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0001-6729-7695}{0000-0001-6729-7695}}\\%
\href{mailto:mail@mitchelloharawild.com}{\nolinkurl{mail@mitchelloharawild.com}}%
}

\address{%
Christophe Dervieux\\
Posit PBC\\%
Paris, France\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0003-4474-2498}{0000-0003-4474-2498}}\\%
\href{mailto:christophe.dervieux@gmail.com}{\nolinkurl{christophe.dervieux@gmail.com}}%
}

\address{%
Heather Turner\\
University of Warwick\\%
Newport, United Kingdom\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-1256-3375}{0000-0002-1256-3375}}\\%
\href{mailto:ht@heatherturner.net}{\nolinkurl{ht@heatherturner.net}}%
}

\address{%
Dianne Cook\\
Monash University\\%
Melbourne, Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-3813-7155}{0000-0002-3813-7155}}\\%
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}%
}
